### 2024-08-25 docs/manuscript.tex
Adding manuscript draft and figures.

\documentclass[fleqn,10pt,doc,onecolumn]{SelfArx}%article} % Document font size and equations flushed left

\usepackage{float}
\usepackage{graphicx}
\usepackage{caption}
\usepackage[hyphens]{url}
\usepackage{hyperref}
\usepackage{xcolor}
\usepackage{amsmath}
\graphicspath{{figures/}}
\setlength{\abovecaptionskip}{15pt plus 3pt minus 2pt}

\newcommand{\beginsupplement}{%
        \setcounter{table}{0}
        \renewcommand{\thetable}{S\arabic{table}}%
        \setcounter{figure}{0}
        \renewcommand{\thefigure}{S\arabic{figure}}%
     }

\definecolor{color1}{RGB}{0,0,90} % Color of the article title and sections
\definecolor{color2}{RGB}{200,200,200} % Color of the boxes behind the abstract and headings
\definecolor{color3}{RGB}{5,5,5} % Color of the boxes behind the abstract and headings

\JournalInfo{.} % Journal information ``Journal, Vol. XXI, No. 1, 1-5, 2015''
\Archive{ } % Additional notes (e.g. copyright, DOI, review/research article)

\PaperTitle{A Bioinformatician, Computer Scientist, and Geneticist lead bioinformatic tool development - which one is better?} % Article
                                                                                                                               % title

\Authors{Paul P. Gardner\textsuperscript{1}*}

\affiliation{\textsuperscript{1}\textit{Department of Biochemistry, University of Otago, Dunedin, New Zealand.}} % Author affiliation
\affiliation{*\textbf{Corresponding author}: paul.gardner@otago.ac.nz} % Corresponding author

\Keywords{} % Keywords - if you don't want any simply remove all the text between the curly brackets
\newcommand{\keywordname}{Keywords} % Defines the keywords heading name

%----------------------------------------------------------------------------------------
%       ABSTRACT
%----------------------------------------------------------------------------------------

% The abstract could benefit from a clearer statement of the research question, methodology, and key findings. It currently mixes multiple ideas without a clear flow.
\Abstract{The development of accurate bioinformatic software tools is
  critical for the effective analysis of complex biological data. This
  study investigates the relationship between the academic department
  affiliations of authors and the accuracy of the bioinformatic tools
  they develop. By analyzing a corpus of previously benchmarked
  bioinformatic software tools, we mapped tools to academic fields of
  corresponding authors and evaluated tool accuracy by field. Our
  results indicate that "Medical Informatics," categorized under
  "Technologies," outperforms other fields in terms of software
  accuracy, with a mean proportion of wins in accuracy rankings higher
  than the null expectation. Conversely, tools developed by authors
  affiliated with "Bioinformatics" and "Engineering" fields produce
  lower accuracy tools. After correcting for multiple testing, no
  result is significant ($p>0.05$). Our findings show a distinct lack
  of an association between academic field and bioinformatic software
  accuracy, highlighting that the development of interdisciplinary
  software applications can be hosted by any aligned department with
  sufficient resources.  }

\begin{document}

\flushbottom % Makes all text pages the same height
\maketitle % Print the title and abstract box
%\tableofcontents % Print the contents section

\thispagestyle{empty} % Removes page numbering from the first page

\section*{Background}

% The introduction lacks a strong hook or clear statement of the problem being addressed. It would be helpful to more explicitly state the importance of understanding the relationship between academic department affiliation and bioinformatics tool accuracy.

Much is made of departmental divisions within academia; These can
denote research and teaching expertise \cite{ben2016s}, influence
hiring decisions, access to funding, publishing and the training of
students recruited for research projects
\cite{bourke1998institutions}. However, interdisciplinary subjects
such as bioinformatics break down the traditional barriers between
departments and subject areas
\cite{Ouzounis:2003,Eddy:2005,hogeweg2011roots}.

Bioinformatics, is an interdisciplinary field that fuses biology,
computer science, and mathematics, and now plays a pivotal role in
modern biological research \cite{Ouzounis:2003}. The development of
bioinformatic tools and software is critical for interpreting complex
biological questions, such as what evolutionary, structural and
functional analyses of genomic, transcriptomic, and proteomic data can
tell us. The field of ``bioinformatics'' can include many overlapping
research fields that include computational biology, biomathematics,
biostatistics, medical informatics and other similar areas.

Bioinformatics began gaining traction with the advent of
high-throughput sequencing technologies, necessitating robust
computational tools to handle the large amounts of data generated
\cite{Ouzounis:2003,hogeweg2011roots}. Departments specializing in
bioinformatics emerged from biology, computer science, and engineering
faculties, each contributing to the field's growth.

Bioinformatics inherently requires a multidisciplinary
approach. Interdisciplinary research allows for the integration of
methods and perspectives from different disciplines, leading to novel
insights and solutions that a single discipline might not achieve
independently \cite{mazzocchi2019scientific}.  This is especially
relevant in bioinformatics, where expertise in biological sciences is
crucial for understanding the data, while computational skills are
needed to develop and implement algorithms, and analyze and interpret
data.

The \textbf{biological and health sciences,} in particular genetics,
biochemistry and molecular biology provide essential domain knowledge,
ensuring that the software tools developed are biologically relevant
and accurate.  Biologists can identify the critical biological
questions and ensure that the computational tools are designed to
address these questions effectively. However, biology departments may
lack the advanced computational expertise required for sophisticated
software development. We have grouped the biological and health fields
into ``\textbf{domain experts}'' in the following analysis.

The \textbf{mathematical, engineering and computational sciences}, or
the ``\textbf{development experts}'' as we have dubbed them here,
contribute significantly by providing expertise in algorithm
development, data structures, and software engineering
principles. These skills are essential for creating efficient,
scalable, and robust bioinformatic tools. Nonetheless, the challenge
for computer science, engineering and mathematics or statistics
departments lies in acquiring a deep understanding of biological
domains necessary to ensure the relevance and accuracy of their
software tools.

Departmental differences might matter in the development of
bioinformatic software tools as they can reflect the the varying
expertise, resources, and perspectives that different academic fields
bring to the table. The development experts may excel in algorithm
efficiency, mathematical modelling, data handling and software engineering. However,
the domain experts may bring a deep understanding of biological
questions, data interpretation and limitations, and possibly better
curation of control datasets. As a result, the success of a tool may
depend more on the integration of diverse expertise rather than the
specific departmental affiliation of its developers. This blending of
skills could neutralize any potential disparities between departments,
leading to comparable outcomes regardless of the department of origin.

As the field of bioinformatics evolves, and the dependence on research
software tools increases due to growing data volumes, this study seeks
to determine whether the academic department of a corresponding author
is associated with the accuracy of the bioinformatic software tools
they develop. Specifically, we investigate whether the presumed
subject expertise, as indicated by the author's departmental
affiliation, has a measurable impact on the accuracy of these tools,
using a benchmarked corpus of bioinformatic software tools to assess
this relationship.

%% \textbf{Engineering}, particularly those focusing on
%% bioengineering and systems biology, bridge the gap between biological
%% sciences and computational techniques. They often have a strong
%% emphasis on interdisciplinary research, which is vital for
%% bioinformatics. These departments can leverage their strengths in both
%% domains to develop integrated solutions.

%% Nick Loman and Mick Watson (2013) argue that a good computational
%% biologist must understand both the biological questions and the
%% computational methods used to address them (Loman \& Watson
%% nbt.2740). This dual expertise is often best nurtured in departments
%% that prioritize interdisciplinary training and collaboration.

%% \textbf{The developer expert, domain expert or interdisciplinary scientist ...}

\begin{figure*}[ht!]
\begin{center}
  \includegraphics[width=0.6\textwidth]{upset-plots.pdf}
\end{center}
  \caption{ The number and intersection of general (\textbf{A}) or
  specific (\textbf{B}) fields that have contributed to bioinformatic
  software tools included in recent benchmark studies.  The number and
  intersection of specific fields.  }
\label{fig:fig1}
\end{figure*}

\section*{Results}

We are interested in the relationship between the accuracy of
bioinformatic software tools and academic fields of study. Using a
published corpus of accuracy rankings for bioinformatic software
tools, we have mapped software tools to academic fields and evaluated
the relationship between software accuracy and academic field.

Previously benchmarked and collated accuracy ranks were obtained from
a published supplement \cite{gardner2024}. We mapped the corresponding
author addresses for published software tools to a standardised list of
specific ``fields of study'' \cite{fields2014}, we further grouped
these into higher level general fields, and broader areas of expertise
based on the address details of each corresponding author. The number
of tools representing each general and specific field is illustrated
in Figure~\ref{fig:fig1}. The majority of bioinformatic
software is produced by corresponding authors who list at least one of
Genetics, Bioinformatics or Computer Science or similar departments as
their primary address (Figure~\ref{fig:fig1}A). For general fields, the
Biological Sciences produce the bulk of software tools, followed by
the Computer Sciences (Figure~\ref{fig:fig1}B).

\begin{figure*}[ht!]
\begin{center}
  \includegraphics[width=0.8\textwidth]{forest-z-Plot.pdf}
\end{center}
\caption{(\textbf{A}) A forest plot, illustrating the mean and 95\%
  confidence intervals of the proportion of times software tools
  published by a given field ``win'' in pairwise
  comparisons. Confidence intervals and the mean was determined using
  a bootstrapping procedure. Within each field the entries have been
  sorted by the mean number of wins. The sample size for each field is
  indicated by the column of numbers on the right of the figure.
  (\textbf{B}) A Z-score was computed for each distribution of
  bootstrap samples for each field. The expected proportion of wins
  for randomly selected groups of tools was used as ``$x$''
  (i.e. null=0.5).   }
\label{fig:fig2}
\end{figure*}

The mean proportion of wins (e.g. if tool `A' outranks tool `B', this
is a win for tool `A') and the corresponding Z-scores give a way to
rank fields of study on the relative accuracy of bioinformatic
software benchmarked between 2005 and 2020 relative to the expected
number of wins for random groupings (i.e. $wins=0.5$). The greater the
proportion of wins, and $(-1)*Z-score$, then the more times tools from
a particular department outperformed competing tools in independent
benchmarks.

The specific and general fields that outperformed other
fields is ``Medical Informatics'' which is a branch of
``Technologies'', the same 16 software tools fall into both
fields. Five of which are different parameter options for the MAFFT
multiple sequence alignment tool \cite{katoh2008recent}.  These have a
mean proportion of wins of 0.70 and 95\% confidence interval of
$0.53-0.85$, which excludes the null of 0.5. The corresponding z-score
is $-1.88$ and $P=0.29$ after correcting for multiple testing.

At the other end of the spectrum is ``Bioinformatics'', ironically
authors that list ``Bioinformatics'' in their address field appear to
write less accurate software for bioinformatic
applications. The mean proportion of wins was 0.43, and 95\%
confidence interval of $0.33-0.53$. The corresponding z-score is
$1.20$ and $P=0.46$ after correcting for multiple
testing.

The general field ``Engineering'' also had a low rank. The mean
proportion of wins was just 0.34, and 95\% confidence interval of
$0.16-0.60$. The corresponding z-score is $1.25$. This general field
is made up of several smaller specific fields that includes
``Bioengineering and biomedical engineering'', ``Computer
engineering'' and ``Electrical and electronics engineering'' which
individually did not have more than ten corresponding software tools,
so were excluded from the more specific analysis.

The remaining general and specific academic fields have confidence
intervals that include the null value of 0.5, and relatively modest
z-scores that range from -0.49 to 0.96. The p-values for each were
greater than $0.05$.

For the highest level field classifications of software development
expert, biological domain expert or interdisciplinary expert each had
similar mean proportions of wins (0.51, 0.49 and 0.46
respectively). The interdisciplinary experts had a lower z-score of
$-0.87$, which corresponds to $P=0.46$ after correcting for multiple
testing.

%followed by ``Molecular Medicine''
%We have deliberately not computed P-values or other measures of
%significance for this study.

\section*{Conclusions and Limitations}

We have tested the assumption that the academic department subject
reflects the quality of research software. Our findings found no
significant association after multiple-testing correction between the
biological domain expert, the software developer expert or the
interdisciplinary scientist with software tool rankings on accuracy
measures. Leading us to conclude that assumptions regarding academic
department and bioinformatic software quality are likely false. The
general and specific research fields yielded similarly negative
results (Figure~\ref{fig:fig2}A).

Our earlier paper found that long-term commitments to updating
software was the leading factor associated with accurate software
tools. This current study complements that finding, showing that not
only do citation metrics (e.g. journal impact factors and author
H-index), tool age and tool speed, but also academic fields of inquiry
are \textbf{not} associated with prediction accuracy.

We focus on software tool accuracy here \cite{weber2019essential}. While speed, usability and
some features of software tools are important, in our opinion the
primary concern for bioinformatic software tools is the accuracy of
the results they produce. As poor predictions may have long-term consequences for
our general research field.

Medical Informatics, under the broader category of "Technologies," is
identified as the top-performing group in developing accurate
bioinformatic software tools. The tools include a number of methods
for structural variation detection, single-cell profiling, long-read
assembly, multiple sequence alignment and are derived from several
different research teams.

Bioinformatics and Engineering ranked lower in terms of software
accuracy. Tools developed by authors who affiliated with
"Bioinformatics" typically had slightly lower accuracy than that of
other fields. However, this was not a statistically significant
finding. In addition (further bad news for this author),
``Biochemistry'' was similarly ranked, again this was not a
statistically significant finding.

This leads us to conclude that an individual's host department are not
reflective of the quality of software that they are capable of
producing ($p>0.05$ in all instances). As a consequence, the academic
department should not be used as a proxy for judging the potential of
software development projects.

%Limitations mentioned in the conclusion could be expanded to better discuss how they may affect the study's findings and their implications.

This study has several limitations. Some benchmarks rank several tool
options, the effects of which may be modest to large, and potentially
non-independent. The number of accuracy metrics used is also diverse,
some of which may be flawed in some circumstances. For example,
``accuracy'' with large class-imbalance \cite{luque2019impact}, or N50
which some commentators have criticised \cite{xie2021pdr}. Some
benchmarks are relatively small resulting in small changes in rank
having a potentially large impact on the proportion of wins. Finally,
the cohort of benchmarks has not been updated to include more recent
results.

There is likely a disconnect between training and the host department
for many bioinformatic tool developers. For example, this author
studied mathematics as an undergraduate, conducted a PhD in
bioinformatics, held postdoctoral fellowships in computer science and
molecular biology departments, spent several years at a genomics
research institute and is now employed in a Biochemistry
Department. The ``Biochemistry'' departmental label is not an accurate
reflection of my training or recent publications. I certainly can't
recall any of the key steps of glycolysis, the citric acid cycle or
the structure of any amino acid or protein.

The last, corresponding author is generally the principal investigator
leading a project. They may have limited involvement in the actual
development of any software tool while they (supposedly) provide
resources for the project and overall direction. However, there is a
significant overlap between the department of a first author, who
would generally be the primary developer of a tool, and the last
author (not tested here). Therefore we expect these results to be
broadly similar if first-author departments were used instead.

%OPen questions: what impact does training have on improving software quality?
%

%We end with echoing the call for more training and a continued 
%\cite{johanson2018software,siepel2019challenges}

\section*{Methods}

\textbf{Pre-registration:} This study's desired sample size, included variables, hypotheses, and
planned analyses were pre-registered on the Open Science Framework
%(https://doi.org/10.17605/OSF.IO/92PTZ)
prior to any unpublished data being collected \cite{gardner2024}.

\textbf{Benchmarking data:} software ranks from previously gathered
benchmarks are publically available \cite{Gardner:2022}, these include data from
68 publications that rank the accuracy of different sets of 498
distinct software tools.

%The methodology section should provide more detail about how the software tools were selected and mapped to academic fields. It’s important to describe how biases or inconsistencies in mapping were addressed.
\textbf{Mapping tools to academic field:} For each software tool, when
available, the corresponding publication(s) were identified and the
addresses for the primary corresponding author were extracted
manually. In cases where an author lists multiple addresses, just the
first two are used. When multiple corresponding authors were listed,
the last corresponding author was selected.

The author department names were mapped to the closest associated
``fields of study'' listed by the National Science Foundation
\cite{fields2014}. We analysed fields at three different hierarchical
levels, firstly the specific fields (e.g. ``genetics'', ``computer
science'', ``bioinformatics'' etc). These are mapped to general fields
within the ``fields of study'' classification (e.g. biological
sciences, computer sciences etc). We have further mapped these to
expertise type, the \textbf{development experts}, \textbf{domain
  experts} and \textbf{interdisciplinary experts}. The development
experts are from fields such as computer science, mathematics and
engineering, they are expected to bring expertise in software
engineering and mathematical modeling of biological problems. The
domain experts are drawn from the biological and health sciences and
are expected to have detailed knowledge of the subject area and be
invested in producing high performance software for their own
research. The interdisciplinary experts are drawn from
interdisciplinary subjects such as bioinformatics and biomathematics,
and from researchers who list both development and domain expertise in
their addresses. We have treated some fields as essentially
synonymous, for example departments of ``Computational Biology'' was
mapped to ``Bioinformatics''.

We restricted all subsequent analyses to fields that contain at least 10
software tools in our benchmark corpus. This mitigated against
potential issues due to small sample sizes.

\textbf{Statistical analysis:} The accuracy data is derived from
benchmarks using a diverse number of metrics that include sensitivity,
specificity, PPV, FDR, error rates, AUROC, MCC and others
\cite{weber2019essential}. The number of tools ranked in any benchmark
ranged from 3 to 50. In order to obtain a representative measure of
accuracy for a field that transcends diverse accuracy measures and
numbers of ranked tools we elected to use ranks and a
bootstrapping strategy.  We randomly sample with replacement sets of
200 tools (from the total of 498), a benchmark that includes each tool
is selected at random and the number of times the tool ``won'' against
another tool is recorded, as is the total number of pairwise
comparisons that were made. These counts of wins and total comparisons
are transferred to the corresponding specific, general and expertise
areas. This is repeated 1,000 times, and allows a way to estimate the
mean proportions of wins allocated to each field, and the 95\%
confidence interval (Figure~\ref{fig:fig2}A). We have also computed a
Z-score for each field using the below equation. This measures the
number of standard deviations the mean number of wins is from the
expected null value of 0.5 for randomly grouped tools
(Figure~\ref{fig:fig2}B).

$z=\frac{x-\mu}{\sigma}$

%\textbf{WRITE UP: Bootstrapping, mean number of wins, confidence intervals, forest plot. }

Where $\mu$ is the mean, $\sigma$ is the standard deviation, $x$ is the raw
value. In this case we set $x=0.5$ as this is the null expectation for
the proportion of wins for randomly grouped sets of tools. For the
purposes of illustration we plot $(-1)*z$ so that the direction is the
same as for the ``proportion of wins'' forest plot
(Figure~\ref{fig:fig1}).

P-values are computed from the absolute value of the z-scores to
evaluate if any field is significantly distinguished from the null
i.e. $P[X > x]$. The P-values are corrected for multiple testing by
controlling the false discovery rate method
\cite{benjamini1995controlling}.

\noindent\textbf{Data and analysis scripts availability}

Add a GitHub Repo...

\noindent\textbf{Funding:} This research was supported by the MBIE data
science platform ``Beyond Prediction: explanatory and transparent data
science'', Genomics Aotearoa, and Marsden
Grants 19-UOO-040 and 20-UOA-282.

\noindent\textbf{Authors' contributions:} this work was conceived by 
PPG, designed by PPG, the analysis was carried out by PPG, interpretation of the results was by PPG,
the manuscript was drafted by  PPG.
All authors revised and edited the final manuscript. All authors approve the submitted version.

%----------------------------------------------------------------------------------------
%       REFERENCE LIST
%----------------------------------------------------------------------------------------
\bibliographystyle{unsrt}
\bibliography{references.bib}

\end{document}

### 2024-08-25 docs/manuscript.tex
Updating manuscript with github link.

@@
 
-\textbf{Pre-registration:} This study's desired sample size, included variables, hypotheses, and
+~~~~\textbf{Pre-registration:} This study's desired sample size, included variables, hypotheses, and
 planned analyses were pre-registered on the Open Science Framework
@@
 levels, firstly the specific fields (e.g. ``genetics'', ``computer
-science'', ``bioinformatics'' etc). These are mapped to general fields
-within the ``fields of study'' classification (e.g. biological
-sciences, computer sciences etc). We have further mapped these to
-expertise type, the \textbf{development experts}, \textbf{domain
-  experts} and \textbf{interdisciplinary experts}. The development
-experts are from fields such as computer science, mathematics and
-engineering, they are expected to bring expertise in software
-engineering and mathematical modeling of biological problems. The
-domain experts are drawn from the biological and health sciences and
-are expected to have detailed knowledge of the subject area and be
-invested in producing high performance software for their own
-research. The interdisciplinary experts are drawn from
-interdisciplinary subjects such as bioinformatics and biomathematics,
-and from researchers who list both development and domain expertise in
-their addresses. We have treated some fields as essentially
-synonymous, for example departments of ``Computational Biology'' was
-mapped to ``Bioinformatics''.
+science'', ``bioinformatics'' etc). These are then mapped to more
+general fields within the ``fields of study'' classification
+(e.g. ``biological sciences'', ``computer sciences'' etc). We have
+further mapped these to three expertise types, the \textbf{development
+  experts}, \textbf{domain experts} and \textbf{interdisciplinary
+  experts}. The development experts are from fields such as computer
+science, mathematics and engineering, they are expected to bring
+development-pertinent expertise in software engineering and
+mathematical modeling of biological problems. The domain experts are
+drawn from the biological and health sciences and are expected to have
+detailed knowledge of the subject area and be invested in producing
+high-performing software for their research. The interdisciplinary
+experts are drawn from interdisciplinary subjects such as
+bioinformatics, biostatistics and biomathematics. Additionally the
+researchers who list both development and domain expertise in their
+addresses (e.g. ``Computer Science'' and ``Genetics''). We have treated some fields as essentially synonymous, for
+example departments of ``Computational Biology'' was mapped to
+``Bioinformatics'', and ``Genomics'' is mapped to ``Genetics''.
 
@@
 areas. This is repeated 1,000 times, and allows a way to estimate the
-mean proportions of wins allocated to each field, and the 95\%
-confidence interval (Figure~\ref{fig:fig2}A). We have also computed a
+mean proportions of wins allocated to each field, and a 95\%
+confidence interval for this value (Figure~\ref{fig:fig2}A). We also compute a
 Z-score for each field using the below equation. This measures the
@@
 
-Add a GitHub Repo...
+The data, scripts, figures and manuscript draft files are availble at the GitHub repository:\\
+\href{https://github.com/ppgardne/departments-software-accuracy}{https://github.com/ppgardne/departments-software-accuracy}
 

### 2024-08-25 bin/plotResults.R

@@ -32,13 +32,11 @@ lines( c(0.5,0.5), c(0,num+1) )
 text(0,num,   "Expertise",      pos=4, col=colours[1])
 text(0,num-3, "General fields",  pos=4, col=colours[2])
 text(0,num-9, "Specific fields", pos=4, col=colours[3])
-#dev.off()
 z      <- -1*(0.5 - meanWins$meanWins)/meanWins$standardDeviation
 p.vals     <- pnorm(abs(z), lower.tail = FALSE)
 p.vals.adj <- p.adjust(p.vals, method = "fdr", n= length(p.vals))
 meanWins <- cbind( cbind(meanWins, z), p.vals.adj)
 write.csv(meanWins, file="fig2-data.tsv")
-#pdf(file=    "../docs/figures/z-scores.pdf", width = 12,  height = 15)
 par(mar = c(4,1,1,9) + .1, cex=2.0, las = 1) #c(bottom, left, top, right). default: c(5, 4, 4, 2) + 0.1
 plot( NA,NA, xlim=c(-1.3,2), ylim=c(1,num), xlab="(-1)*Z-score (null=0.5)", ylab="", main="", yaxt = "n" )
 for(i in 1:num){

### 2024-08-26 docs/manuscript.tex
Revised manuscript for clarity.

@@
 % The abstract could benefit from a clearer statement of the research question, methodology, and key findings. It currently mixes multiple ideas without a clear flow.
-\Abstract{The development of accurate bioinformatic software tools is
-  critical for the effective analysis of complex biological data. This
-  study investigates the relationship between the academic department
+\Abstract{
+  The development of accurate bioinformatic software tools is crucial
+  for the effective analysis of complex biological data. This study
+  examines the relationship between the academic department
   affiliations of authors and the accuracy of the bioinformatic tools
   they develop. By analyzing a corpus of previously benchmarked
-  bioinformatic software tools, we mapped tools to academic fields of
-  corresponding authors and evaluated tool accuracy by field. Our
-  results indicate that "Medical Informatics," categorized under
-  "Technologies," outperforms other fields in terms of software
-  accuracy, with a mean proportion of wins in accuracy rankings higher
-  than the null expectation. Conversely, tools developed by authors
-  affiliated with "Bioinformatics" and "Engineering" fields produce
-  lower accuracy tools. After correcting for multiple testing, no
-  result is significant ($p>0.05$). Our findings show a distinct lack
-  of an association between academic field and bioinformatic software
-  accuracy, highlighting that the development of interdisciplinary
-  software applications can be hosted by any aligned department with
-  sufficient resources.  }
+  bioinformatic software tools, we mapped bioinformatic tools to the
+  academic fields of the corresponding authors and evaluated tool
+  accuracy by field. Our results suggest that "Medical Informatics,"
+  under the broader category of "Technologies," outperforms other
+  fields in software accuracy, with a mean proportion of wins in
+  accuracy rankings exceeding the null expectation. In contrast, tools
+  developed by authors affiliated with "Bioinformatics" and
+  "Engineering" fields tend to be less accurate. However, after
+  correcting for multiple testing, none of the results are
+  statistically significant ($p>0.05$). Our findings reveal no
+  association between academic field and bioinformatic
+  software accuracy, suggesting that the development of
+  interdisciplinary software applications can be effectively
+  undertaken by any department with sufficient resources and training.
+}
 
@@
 
-Bioinformatics began gaining traction with the advent of
-high-throughput sequencing technologies, necessitating robust
-computational tools to handle the large amounts of data generated
-\cite{Ouzounis:2003,hogeweg2011roots}. Departments specializing in
-bioinformatics emerged from biology, computer science, and engineering
-faculties, each contributing to the field's growth.
+Bioinformatics gained significant traction with the advent of
+high-throughput sequencing technologies, which created a need for
+robust computational tools to manage the large volumes of data
+generated \cite{Ouzounis:2003,hogeweg2011roots}. In response,
+departments specializing in bioinformatics began to emerge from
+biology, computer science, and engineering faculties, each bringing
+unique contributions to the field's development and expansion.
 
-Bioinformatics inherently requires a multidisciplinary
-approach. Interdisciplinary research allows for the integration of
-methods and perspectives from different disciplines, leading to novel
-insights and solutions that a single discipline might not achieve
-independently \cite{mazzocchi2019scientific}.  This is especially
-relevant in bioinformatics, where expertise in biological sciences is
-crucial for understanding the data, while computational skills are
-needed to develop and implement algorithms, and analyze and interpret
-data.
+The interdisciplinarity of bioinformatics enables the integration of
+methods and perspectives from various disciplines, fostering novel
+insights and solutions that might not be possible within a single
+field \cite{mazzocchi2019scientific}. This requires a deep
+understanding of biological sciences for interpreting
+data, and strong computational skills are required to develop
+algorithms and analyze the results at scale.
 
@@
 biochemistry and molecular biology provide essential domain knowledge,
-ensuring that the software tools developed are biologically relevant
-and accurate.  Biologists can identify the critical biological
-questions and ensure that the computational tools are designed to
-address these questions effectively. However, biology departments may
-lack the advanced computational expertise required for sophisticated
-software development. We have grouped the biological and health fields
-into ``\textbf{domain experts}'' in the following analysis.
+ensuring that the developed software tools are both biologically relevant
+and accurate.  Biologists are well-positioned to identify critical biological
+questions and ensure that computational tools are designed to
+effectively address these. However, biology departments may
+lack the advanced computational expertise required for developing sophisticated
+software. In our analysis, we have grouped the biological and health fields
+under the category of ``\textbf{domain experts}''.
 
-The \textbf{mathematical, engineering and computational sciences}, or
-the ``\textbf{development experts}'' as we have dubbed them here,
-contribute significantly by providing expertise in algorithm
-development, data structures, and software engineering
+The \textbf{mathematical, engineering and computational sciences},
+referred to here as the ``\textbf{development experts}'', contribute
+significantly by bringing expertise in algorithm development,
+mathematical modelling, statistics, and software engineering
 principles. These skills are essential for creating efficient,
-scalable, and robust bioinformatic tools. Nonetheless, the challenge
-for computer science, engineering and mathematics or statistics
-departments lies in acquiring a deep understanding of biological
-domains necessary to ensure the relevance and accuracy of their
-software tools.
+scalable, and robust bioinformatic tools.  However, the challenge for
+development experts is gaining a deep understanding of biological domains,
+which is necessary to ensure software tools are both relevant
+and accurate.
 
-Departmental differences might matter in the development of
-bioinformatic software tools as they can reflect the the varying
+Departmental differences can influence the development of
+bioinformatic software tools because they reflect the varying
 expertise, resources, and perspectives that different academic fields
-bring to the table. The development experts may excel in algorithm
-efficiency, mathematical modelling, data handling and software engineering. However,
-the domain experts may bring a deep understanding of biological
-questions, data interpretation and limitations, and possibly better
-curation of control datasets. As a result, the success of a tool may
-depend more on the integration of diverse expertise rather than the
-specific departmental affiliation of its developers. This blending of
-skills could neutralize any potential disparities between departments,
-leading to comparable outcomes regardless of the department of origin.
+offer. Development experts may excel in areas such as algorithm
+efficiency, mathematical modeling, data handling, and software
+engineering, while domain experts may bring a deep understanding of
+biological questions, data interpretation, limitations, and
+potentially better curation of control datasets. Consequently, the
+success of a tool may rely more on the integration of diverse
+forms of expertise rather than the specific departmental affiliation
+of its developers. This blending of skills can mitigate potential
+disparities between departments, resulting in comparable outcomes
+regardless of the department of origin.
 
-As the field of bioinformatics evolves, and the dependence on research
-software tools increases due to growing data volumes, this study seeks
-to determine whether the academic department of a corresponding author
-is associated with the accuracy of the bioinformatic software tools
-they develop. Specifically, we investigate whether the presumed
-subject expertise, as indicated by the author's departmental
-affiliation, has a measurable impact on the accuracy of these tools,
-using a benchmarked corpus of bioinformatic software tools to assess
-this relationship.
-
-%% \textbf{Engineering}, particularly those focusing on
-%% bioengineering and systems biology, bridge the gap between biological
-%% sciences and computational techniques. They often have a strong
-%% emphasis on interdisciplinary research, which is vital for
-%% bioinformatics. These departments can leverage their strengths in both
-%% domains to develop integrated solutions.
-
-%% Nick Loman and Mick Watson (2013) argue that a good computational
-%% biologist must understand both the biological questions and the
-%% computational methods used to address them (Loman \& Watson
-%% nbt.2740). This dual expertise is often best nurtured in departments
-%% that prioritize interdisciplinary training and collaboration.
-
-%% \textbf{The developer expert, domain expert or interdisciplinary scientist ...}
+As the field of bioinformatics evolves and reliance on research
+software tools grows with increasing data volumes, this study aims to
+determine whether the academic department of a corresponding author
+influences the accuracy of the bioinformatic software tools they
+develop. Specifically, we examine whether the presumed subject
+expertise, reflected in the author's departmental affiliation, has a
+measurable impact on tool accuracy, utilizing a benchmarked corpus of
+bioinformatic software tools to evaluate this relationship.
 
@@
 
-We are interested in the relationship between the accuracy of
-bioinformatic software tools and academic fields of study. Using a
-published corpus of accuracy rankings for bioinformatic software
-tools, we have mapped software tools to academic fields and evaluated
-the relationship between software accuracy and academic field.
+We are interested in exploring the relationship between the accuracy
+of bioinformatic software tools and the academic fields of study of
+their developers. Using a published corpus of accuracy rankings
+for bioinformatic software tools, we mapped these tools to their
+respective academic fields and evaluated how software accuracy
+correlates with the developers' academic affiliations.
 
-Previously benchmarked and collated accuracy ranks were obtained from
-a published supplement \cite{gardner2024}. We mapped the corresponding
-author addresses for published software tools to a standardised list of
-specific ``fields of study'' \cite{fields2014}, we further grouped
-these into higher level general fields, and broader areas of expertise
-based on the address details of each corresponding author. The number
-of tools representing each general and specific field is illustrated
-in Figure~\ref{fig:fig1}. The majority of bioinformatic
-software is produced by corresponding authors who list at least one of
-Genetics, Bioinformatics or Computer Science or similar departments as
-their primary address (Figure~\ref{fig:fig1}A). For general fields, the
-Biological Sciences produce the bulk of software tools, followed by
+We obtained previously benchmarked and collated accuracy rankings from
+a published supplement \cite{gardner2024}. Using these rankings, we
+mapped the addresses of corresponding authors for each published
+software tool to a standardized list of specific ``fields of study''
+\cite{fields2014}. These fields were then grouped into higher-level
+general fields and broader areas of expertise based on the address
+details of each corresponding author. Figure~\ref{fig:fig1} illustrates
+the number of tools representing each general and specific field. Most
+bioinformatic software is developed by corresponding authors who list
+Genetics, Bioinformatics, Computer Science, or similar departments as
+their primary address (Figure~\ref{fig:fig1}A). Among the general fields,
+the Biological Sciences produce the most software tools, followed by
 the Computer Sciences (Figure~\ref{fig:fig1}B).
@@
   for randomly selected groups of tools was used as ``$x$''
-  (i.e. null=0.5).   }
+  (i.e. null=0.5).}
 \label{fig:fig2}
@@
 
-The mean proportion of wins (e.g. if tool `A' outranks tool `B', this
-is a win for tool `A') and the corresponding Z-scores give a way to
-rank fields of study on the relative accuracy of bioinformatic
-software benchmarked between 2005 and 2020 relative to the expected
-number of wins for random groupings (i.e. $wins=0.5$). The greater the
-proportion of wins, and $(-1)*Z-score$, then the more times tools from
-a particular department outperformed competing tools in independent
-benchmarks.
+The mean proportion of wins (e.g., if tool `A' outranks tool `B', this
+counts as a win for tool `A') and the corresponding Z-scores provide a
+method to rank fields of study based on the relative accuracy of
+bioinformatic software benchmarked between 2005 and 2020, compared to
+the expected number of wins for random groupings (i.e., $wins =
+0.5$). A higher proportion of wins and a greater $(-1) * Z-score$
+indicate that tools from a particular department more frequently
+outperformed competing tools in independent benchmarks.
 
-The specific and general fields that outperformed other
-fields is ``Medical Informatics'' which is a branch of
-``Technologies'', the same 16 software tools fall into both
-fields. Five of which are different parameter options for the MAFFT
-multiple sequence alignment tool \cite{katoh2008recent}.  These have a
-mean proportion of wins of 0.70 and 95\% confidence interval of
-$0.53-0.85$, which excludes the null of 0.5. The corresponding z-score
-is $-1.88$ and $P=0.29$ after correcting for multiple testing.
+The specific and general field that outperformed all others is
+``Medical Informatics'', a branch of ``Technologies'', with 16
+software tools categorized under both fields. Among these, five are
+different parameter options for the MAFFT multiple sequence alignment
+tool \cite{katoh2008recent}.  These Medical Informatic tools have a mean proportion of wins of
+0.70 and 95\% confidence interval of $0.53-0.85$, which excludes the
+null value of 0.5. The corresponding Z-score is $-1.88$, and after correcting for multiple testing, the P-value is $0.29$.
 
-At the other end of the spectrum is ``Bioinformatics'', ironically
-authors that list ``Bioinformatics'' in their address field appear to
-write less accurate software for bioinformatic
-applications. The mean proportion of wins was 0.43, and 95\%
-confidence interval of $0.33-0.53$. The corresponding z-score is
-$1.20$ and $P=0.46$ after correcting for multiple
-testing.
+At the other end of the spectrum is ``Bioinformatics''. Ironically,
+authors that list ``Bioinformatics'' in their address field tend to
+produce less accurate software for bioinformatic applications. The
+mean proportion of wins was 0.43, with a 95\% confidence interval of
+$0.33-0.53$. The corresponding Z-score is $1.20$ and after correcting
+for multiple testing the P-value is $0.46$.
 
-The general field ``Engineering'' also had a low rank. The mean
-proportion of wins was just 0.34, and 95\% confidence interval of
-$0.16-0.60$. The corresponding z-score is $1.25$. This general field
-is made up of several smaller specific fields that includes
-``Bioengineering and biomedical engineering'', ``Computer
-engineering'' and ``Electrical and electronics engineering'' which
-individually did not have more than ten corresponding software tools,
-so were excluded from the more specific analysis.
+The general field of ``Engineering'' also ranked low, with a mean
+proportion of wins of just 0.34 and a 95\% confidence interval of
+$0.16-0.60$. The corresponding Z-score is $1.25$. This general field
+comprises several smaller specific fields, such as "Bioengineering and
+Biomedical Engineering," "Computer Engineering," and "Electrical and
+Electronics Engineering." However, since none of these specific fields
+individually had more than ten corresponding software tools, they were
+excluded from the more detailed analysis.
 
@@
 intervals that include the null value of 0.5, and relatively modest
-z-scores that range from -0.49 to 0.96. The p-values for each were
+Z-scores that range from -0.49 to 0.96. The P-values for each were
 greater than $0.05$.
@@
 similar mean proportions of wins (0.51, 0.49 and 0.46
-respectively). The interdisciplinary experts had a lower z-score of
+respectively). The interdisciplinary experts had a lower Z-score of
 $-0.87$, which corresponds to $P=0.46$ after correcting for multiple
@@
 
-We have tested the assumption that the academic department subject
-reflects the quality of research software. Our findings found no
-significant association after multiple-testing correction between the
-biological domain expert, the software developer expert or the
-interdisciplinary scientist with software tool rankings on accuracy
-measures. Leading us to conclude that assumptions regarding academic
-department and bioinformatic software quality are likely false. The
-general and specific research fields yielded similarly negative
-results (Figure~\ref{fig:fig2}A).
+We tested the assumption that the speciality of academic departments
+reflect the quality of the research software they produce. Our
+findings showed no significant association, after correcting for
+multiple-testing, between the type of academic expertise -- whether
+biological domain expert, software developer, or interdisciplinary
+scientist -- and the accuracy rankings of bioinformatic software
+tools. This leads us to conclude that assumptions linking academic
+department affiliation to the quality of bioinformatic software are
+likely incorrect. Similarly, both general and specific research fields
+showed no significant associations (Figure~\ref{fig:fig2}).
 
-Our earlier paper found that long-term commitments to updating
-software was the leading factor associated with accurate software
-tools. This current study complements that finding, showing that not
-only do citation metrics (e.g. journal impact factors and author
-H-index), tool age and tool speed, but also academic fields of inquiry
-are \textbf{not} associated with prediction accuracy.
+Our earlier paper found that a long-term commitment to keeping
+software updated was the primary factor associated with accurate
+software tools. This current study complements that finding by demonstrating 
+that citation metrics (e.g. journal impact factors and
+author H-index), tool age, tool speed, and now academic fields of
+inquiry are \textbf{not} associated with software accuracy.
 
@@
 
-%Limitations mentioned in the conclusion could be expanded to better discuss how they may affect the study's findings and their implications.
+\textbf{Limitations:} Some benchmarks rank multiple tool options,
+which can introduce modest to large effects as these options may not
+be independent. Additionally, the accuracy metrics used are diverse,
+and some may be flawed under certain conditions; for example,
+``accuracy'' can be misleading with large class-imbalances
+\cite{luque2019impact}, and the N50 metric for sequence assembly has
+been criticized by some commentators \cite{xie2021pdr}. Moreover, some
+benchmarks are relatively small, so even small changes in rank can
+have a substantial impact on the proportion of wins for intermediate
+rankings. Lastly, the cohort of benchmarks has not been updated to
+include more recent results.
 
-This study has several limitations. Some benchmarks rank several tool
-options, the effects of which may be modest to large, and potentially
-non-independent. The number of accuracy metrics used is also diverse,
-some of which may be flawed in some circumstances. For example,
-``accuracy'' with large class-imbalance \cite{luque2019impact}, or N50
-which some commentators have criticised \cite{xie2021pdr}. Some
-benchmarks are relatively small resulting in small changes in rank
-having a potentially large impact on the proportion of wins. Finally,
-the cohort of benchmarks has not been updated to include more recent
-results.
+There is often a disconnect between a bioinformatic tool developer's
+training and their host department. For example, the author of this
+study completed an undergraduate degree in mathematics, a PhD in
+bioinformatics, postdoctoral fellowships in computer science and
+molecular biology departments, worked for several years at a genomics
+research institute, and is currently employed in a Biochemistry
+Department. The ``Biochemistry'' departmental label does not accurately
+reflect the author's training or recent publications. In fact, the
+author admits to not recalling key details of glycolysis, the citric
+acid cycle, or the structures of amino acids and proteins.
 
-There is likely a disconnect between training and the host department
-for many bioinformatic tool developers. For example, this author
-studied mathematics as an undergraduate, conducted a PhD in
-bioinformatics, held postdoctoral fellowships in computer science and
-molecular biology departments, spent several years at a genomics
-research institute and is now employed in a Biochemistry
-Department. The ``Biochemistry'' departmental label is not an accurate
-reflection of my training or recent publications. I certainly can't
-recall any of the key steps of glycolysis, the citric acid cycle or
-the structure of any amino acid or protein.
+The last, or corresponding, author is typically the principal
+investigator who leads a project. They may have limited direct
+involvement in the development of any software tool, while
+their role is mainly to provide resources and overall direction for
+the project. However, there is likely to be a significant overlap between the
+department of the first author, who is usually the primary developer
+of the tool, and the department of the last author (though this was
+not tested in this study). Therefore, we expect that the results will
+be broadly similar if first-author departments were analyzed instead.
 
-The last, corresponding author is generally the principal investigator
-leading a project. They may have limited involvement in the actual
-development of any software tool while they (supposedly) provide
-resources for the project and overall direction. However, there is a
-significant overlap between the department of a first author, who
-would generally be the primary developer of a tool, and the last
-author (not tested here). Therefore we expect these results to be
-broadly similar if first-author departments were used instead.
-
-%OPen questions: what impact does training have on improving software quality?
-%
-
-%We end with echoing the call for more training and a continued 
-%\cite{johanson2018software,siepel2019challenges}
+%Open questions: what impact does training have on improving software quality?
+%We could end with echoing the call for more training and a continued 
+%e.g. \cite{johanson2018software,siepel2019challenges}
 
@@
 It’s important to describe how biases or inconsistencies in mapping were addressed.
-\textbf{Mapping tools to academic field:} For each software tool, when
-available, the corresponding publication(s) were identified and the
-addresses for the primary corresponding author were extracted
-manually. In cases where an author lists multiple addresses, just the
-first two are used. When multiple corresponding authors were listed,
-the last corresponding author was selected.
+\textbf{Mapping tools to academic field:} For each software tool, the
+corresponding publication(s) were identified, and the addresses of the
+primary corresponding author were manually extracted when
+available. If an author listed multiple addresses, only the first two
+were used. In cases with multiple corresponding authors, the last
+corresponding author was chosen.
 
-The author department names were mapped to the closest associated
-``fields of study'' listed by the National Science Foundation
-\cite{fields2014}. We analysed fields at three different hierarchical
-levels, firstly the specific fields (e.g. ``genetics'', ``computer
-science'', ``bioinformatics'' etc). These are then mapped to more
-general fields within the ``fields of study'' classification
-(e.g. ``biological sciences'', ``computer sciences'' etc). We have
-further mapped these to three expertise types, the \textbf{development
-  experts}, \textbf{domain experts} and \textbf{interdisciplinary
-  experts}. The development experts are from fields such as computer
-science, mathematics and engineering, they are expected to bring
-development-pertinent expertise in software engineering and
-mathematical modeling of biological problems. The domain experts are
-drawn from the biological and health sciences and are expected to have
-detailed knowledge of the subject area and be invested in producing
-high-performing software for their research. The interdisciplinary
-experts are drawn from interdisciplinary subjects such as
-bioinformatics, biostatistics and biomathematics. Additionally the
-researchers who list both development and domain expertise in their
-addresses (e.g. ``Computer Science'' and ``Genetics''). We have treated some fields as essentially synonymous, for
-example departments of ``Computational Biology'' was mapped to
-``Bioinformatics'', and ``Genomics'' is mapped to ``Genetics''.
+The department names of the authors were mapped to the closest
+associated ``fields of study'' as defined by the National Science
+Foundation \cite{fields2014}. We analysed these fields at three
+hierarchical levels: first, specific fields (e.g. ``genetics'',
+``computer science'', ``bioinformatics'' etc), which were then mapped
+to broader general fields (e.g. ``biological sciences'', ``computer
+sciences'' etc). Thirdly, we categorized them into three types of
+expertise: \textbf{development experts}, \textbf{domain experts} and
+\textbf{interdisciplinary experts}.  Development experts, from fields
+such as computer science, mathematics, and engineering, are expected
+to bring relevant expertise in software engineering and the
+mathematical modeling of biological problems. Domain experts, from the
+biological and health sciences, are anticipated to possess detailed
+knowledge of their subject area and to be invested in producing
+high-performing software for their research needs. Interdisciplinary
+experts come from fields such as bioinformatics, biostatistics, and
+biomathematics, and also include researchers who list both development
+and domain expertise (e.g. ``Computer Science'' and ``Genetics''). We
+have treated some fields as synonymous; for example, ``Computational
+Biology'' was mapped to ``Bioinformatics'', and ``Genomics'' is mapped
+to ``Genetics''.
 
@@
 ranged from 3 to 50. In order to obtain a representative measure of
-accuracy for a field that transcends diverse accuracy measures and
-numbers of ranked tools we elected to use ranks and a
-bootstrapping strategy.  We randomly sample with replacement sets of
-200 tools (from the total of 498), a benchmark that includes each tool
-is selected at random and the number of times the tool ``won'' against
-another tool is recorded, as is the total number of pairwise
-comparisons that were made. These counts of wins and total comparisons
-are transferred to the corresponding specific, general and expertise
-areas. This is repeated 1,000 times, and allows a way to estimate the
-mean proportions of wins allocated to each field, and a 95\%
-confidence interval for this value (Figure~\ref{fig:fig2}A). We also compute a
-Z-score for each field using the below equation. This measures the
-number of standard deviations the mean number of wins is from the
-expected null value of 0.5 for randomly grouped tools
+accuracy for a field that accounts for the diversity in accuracy measures and
+number of ranked tools, we employed a rank-based and
+bootstrapping strategy.
+We randomly sampled, with replacement, sets of 200 tools from the total of 498 tools.
+For each tool, a corresponding benchmark was selected at random, and the number of times the tool ``won'' against another tool was recorded, along with the total number of pairwise comparisons made.
+These counts of wins and total comparisons were then assigned to the corresponding specific, general, and expertise areas.
+This process was repeated 1,000 times to estimate the mean proportions of wins for each field, along with a 95\% confidence interval for these values (Figure~\ref{fig:fig2}A).
+Additionally, we calculated a Z-score for each field to determine the number of standard deviations the mean number of wins deviates from the expected null value of 0.5 for randomly grouped tools
 (Figure~\ref{fig:fig2}B).
 
-$z=\frac{x-\mu}{\sigma}$
-
-%\textbf{WRITE UP: Bootstrapping, mean number of wins, confidence intervals, forest plot. }
+$Z-score=\frac{x-\mu}{\sigma}$
 
@@
 
-P-values are computed from the absolute value of the z-scores to
+P-values are computed from the absolute value of the Z-scores to
 evaluate if any field is significantly distinguished from the null

### 2024-08-26 docs/manuscript.tex
Further manuscript edits.

@@
 % The abstract could benefit from a clearer statement of the research question, methodology, and key findings. It currently mixes multiple ideas without a clear flow.
-\Abstract{
-  The development of accurate bioinformatic software tools is crucial
-  for the effective analysis of complex biological data. This study
-  examines the relationship between the academic department
+\Abstract{ The development of accurate bioinformatic software tools is
+  crucial for the effective analysis of complex biological data. This
+  study examines the relationship between the academic department
   affiliations of authors and the accuracy of the bioinformatic tools
@@
   academic fields of the corresponding authors and evaluated tool
-  accuracy by field. Our results suggest that "Medical Informatics,"
-  under the broader category of "Technologies," outperforms other
-  fields in software accuracy, with a mean proportion of wins in
-  accuracy rankings exceeding the null expectation. In contrast, tools
-  developed by authors affiliated with "Bioinformatics" and
-  "Engineering" fields tend to be less accurate. However, after
-  correcting for multiple testing, none of the results are
-  statistically significant ($p>0.05$). Our findings reveal no
-  association between academic field and bioinformatic
-  software accuracy, suggesting that the development of
-  interdisciplinary software applications can be effectively
-  undertaken by any department with sufficient resources and training.
-}
+  accuracy by field. Our results suggest that ``Medical Informatics''
+  outperforms all other fields in bioinformatic software accuracy,
+  with a mean proportion of wins in accuracy rankings exceeding the
+  null expectation. In contrast, tools developed by authors affiliated
+  with "Bioinformatics" and "Engineering" fields tend to be less
+  accurate. However, after correcting for multiple testing, none of
+  the results are statistically significant ($p>0.05$). Our findings
+  reveal no strong association between academic field and
+  bioinformatic software accuracy. These findings suggest that the
+  development of interdisciplinary software applications can be
+  effectively undertaken by any department with sufficient resources
+  and training.  }
 
@@
 
-The interdisciplinarity of bioinformatics enables the integration of
-methods and perspectives from various disciplines, fostering novel
-insights and solutions that might not be possible within a single
-field \cite{mazzocchi2019scientific}. This requires a deep
-understanding of biological sciences for interpreting
-data, and strong computational skills are required to develop
-algorithms and analyze the results at scale.
+Bioinformatics integrates methods from various disciplines, fostering
+novel insights that are unattainable within a single field
+\cite{mazzocchi2019scientific}. This requires a deep understanding of
+biological sciences for interpreting data, and strong computational
+skills are required to develop algorithms and analyze the results at
+scale.
 
-The \textbf{biological and health sciences,} in particular genetics,
-biochemistry and molecular biology provide essential domain knowledge,
-ensuring that the developed software tools are both biologically relevant
-and accurate.  Biologists are well-positioned to identify critical biological
-questions and ensure that computational tools are designed to
-effectively address these. However, biology departments may
-lack the advanced computational expertise required for developing sophisticated
-software. In our analysis, we have grouped the biological and health fields
-under the category of ``\textbf{domain experts}''.
+Experts in \textbf{biological and health sciences}, such as genetics
+and molecular biology, can ensure that software tools are biologically
+relevant and accurate by identifying critical questions and guiding
+computational tool development. However, biology departments may lack
+the advanced computational expertise required for developing
+sophisticated software. In our analysis, we have grouped the
+biological and health fields under the category of ``\textbf{domain
+  experts}''.
 
@@
 
-As the field of bioinformatics evolves and reliance on research
-software tools grows with increasing data volumes, this study aims to
-determine whether the academic department of a corresponding author
-influences the accuracy of the bioinformatic software tools they
-develop. Specifically, we examine whether the presumed subject
-expertise, reflected in the author's departmental affiliation, has a
-measurable impact on tool accuracy, utilizing a benchmarked corpus of
-bioinformatic software tools to evaluate this relationship.
+The primary objective of this study is to investigate whether the
+academic department affiliation of a corresponding author is
+associated with the accuracy of the bioinformatic software tools they
+develop. We aim to determine if tools developed by authors from
+certain academic fields, such as ``Bioinformatics'', ``Computer
+Science'' or ``Genetics'' show higher or lower accuracy compared to
+those from other fields. To achieve this, we analyzed a benchmarked
+corpus of bioinformatic software tools and evaluated their accuracy
+based on the academic affiliations of their developers.
 
@@
 
-We obtained previously benchmarked and collated accuracy rankings from
-a published supplement \cite{gardner2024}. Using these rankings, we
-mapped the addresses of corresponding authors for each published
-software tool to a standardized list of specific ``fields of study''
-\cite{fields2014}. These fields were then grouped into higher-level
-general fields and broader areas of expertise based on the address
-details of each corresponding author. Figure~\ref{fig:fig1} illustrates
-the number of tools representing each general and specific field. Most
+We used a published supplement \cite{gardner2024} to obtain previously
+benchmarked accuracy rankings and mapped corresponding authors'
+addresses to a standardized list of specific 'fields of study'
+\cite{fields2014}. These were further grouped into broader general
+fields and areas of expertise. Figure~\ref{fig:fig1} illustrates the
+number of tools representing each general and specific field ($N\ge 10$). Most
 bioinformatic software is developed by corresponding authors who list
 Genetics, Bioinformatics, Computer Science, or similar departments as
-their primary address (Figure~\ref{fig:fig1}A). Among the general fields,
-the Biological Sciences produce the most software tools, followed by
-the Computer Sciences (Figure~\ref{fig:fig1}B).
+their primary address (Figure~\ref{fig:fig1}A). Among the general
+fields, the Biological Sciences produce the most software tools,
+followed by the Computer Sciences (Figure~\ref{fig:fig1}B).
 
@@
 We tested the assumption that the speciality of academic departments
-reflect the quality of the research software they produce. Our
-findings showed no significant association, after correcting for
-multiple-testing, between the type of academic expertise -- whether
-biological domain expert, software developer, or interdisciplinary
-scientist -- and the accuracy rankings of bioinformatic software
-tools. This leads us to conclude that assumptions linking academic
-department affiliation to the quality of bioinformatic software are
-likely incorrect. Similarly, both general and specific research fields
-showed no significant associations (Figure~\ref{fig:fig2}).
+reflect the quality of the research software they produce.  After
+correcting for multiple testing, we found no significant association
+between academic expertise type and the accuracy of bioinformatic
+tools, suggesting that academic department affiliation does not
+correlate with software quality. Similarly, both general and specific
+research fields showed no significant associations
+(Figure~\ref{fig:fig2}).
 
@@
 
-There is often a disconnect between a bioinformatic tool developer's
-training and their host department. For example, the author of this
-study completed an undergraduate degree in mathematics, a PhD in
-bioinformatics, postdoctoral fellowships in computer science and
-molecular biology departments, worked for several years at a genomics
-research institute, and is currently employed in a Biochemistry
-Department. The ``Biochemistry'' departmental label does not accurately
-reflect the author's training or recent publications. In fact, the
-author admits to not recalling key details of glycolysis, the citric
-acid cycle, or the structures of amino acids and proteins.
+There is often a disconnect between a developer’s training and their
+departmental affiliation, as evidenced by this author whose diverse
+academic and professional background spans mathematics,
+bioinformatics, computer science, molecular biology, and genomics, yet
+currently lists their affiliation with a Biochemistry Department. The
+``Biochemistry'' departmental label does not accurately reflect the
+author's training or recent publications. In fact, the author admits
+to not recalling key details of glycolysis, the citric acid cycle, or
+the structures of amino acids and proteins.
 
@@
 be broadly similar if first-author departments were analyzed instead.
+
+In conclusion, this study does not find strong evidence that the
+academic department affiliation of authors is associated with the
+accuracy of bioinformatic software tools. This challenges the
+assumption that departmental expertise correlates with
+software quality. Future research could explore other factors, such
+as the nature of interdisciplinary collaborations or the training
+of developers, to better understand what factors contribute to
+the development of high-quality bioinformatic tools.
 

### 2024-08-29 docs/manuscript.tex
Adding line numbers.

@@
 \usepackage{amsmath}
+\usepackage[switch,pagewise]{lineno}
+\linenumbers
+
 \graphicspath{{figures/}}
@@
 
-\Keywords{} % Keywords - if you don't want any simply remove all the text between the curly brackets
+\Keywords{
+Bioinformatics,
+Software Tool Development, 
+Interdisciplinary Research, 
+Departmental Affiliation, 
+Software Accuracy, 
+Medical Informatics, 
+Academic Expertise, 
+Benchmarking,
+Computational Biology
+} % Keywords - if you don't want any simply remove all the text between the curly brackets
 \newcommand{\keywordname}{Keywords} % Defines the keywords heading name
@@
 \thispagestyle{empty} % Removes page numbering from the first page
+
+\section*{Key Points}
+
+\begin{itemize}
+\item The study investigates the relationship between academic department affiliations of authors and the accuracy of bioinformatic software tools they develop.
+\item Results indicate no statistically significant association between departmental affiliation and the accuracy of bioinformatic tools, challenging the assumption that departmental expertise predicts software quality.
+\item Tools from the "Medical Informatics" field showed a higher mean proportion of wins, while those from "Bioinformatics" and "Engineering" fields ranked lower, though these differences were not statistically significant after correcting for multiple-testing.
+\item The findings suggest that allocating sufficient resources for long term software development are more critical to accurate bioinformatic tools than departmental expertise.
+\end{itemize}
 
@@
 
+%Generate a doi:
+%https://docs.github.com/en/repositories/archiving-a-github-repository/referencing-and-citing-content
+%https://zenodo.org/account/settings/github/
+
 \noindent\textbf{Funding:} This research was supported by the MBIE data

### 2024-09-19 docs/manuscript.tex
Edited manuscript text for clarity and brevity.

--- 2024-08-26 docs/manuscript.tex
@@
 
-\PaperTitle{A Bioinformatician, Computer Scientist, and Geneticist lead bioinformatic tool development - which one is better?} % Article
-                                                                                                                               % title
+\PaperTitle{A Bioinformatician, Computer Scientist, and Geneticist lead bioinformatic tool development - which one is better?}
 
@@
 %----------------------------------------------------------------------------------------
-
-% The abstract could benefit from a clearer statement of the research question, methodology, and key findings. It currently mixes multiple ideas without a clear flow.
-\Abstract{ The development of accurate bioinformatic software tools is
+\Abstract{The development of accurate bioinformatic software tools is
   crucial for the effective analysis of complex biological data. This
@@
   with "Bioinformatics" and "Engineering" fields tend to be less
-  accurate. However, after correcting for multiple testing, none of
-  the results are statistically significant ($p>0.05$). Our findings
+  accurate. However, after correcting for multiple testing, no
+  result is statistically significant ($p>0.05$). Our findings
   reveal no strong association between academic field and
@@
   effectively undertaken by any department with sufficient resources
-  and training.  }
+  and training.}
 
@@
 
-% The introduction lacks a strong hook or clear statement of the problem being addressed.
-It would be helpful to more explicitly state the importance of understanding the relationship between academic department affiliation and bioinformatics tool accuracy.
-
-Much is made of departmental divisions within academia; These can
-denote research and teaching expertise \cite{ben2016s}, influence
-hiring decisions, access to funding, publishing and the training of
-students recruited for research projects
-\cite{bourke1998institutions}. However, interdisciplinary subjects
-such as bioinformatics break down the traditional barriers between
-departments and subject areas
+Departmental divisions in academia signify research expertise,
+influence hiring decisions, and impact access to funding, publishing
+opportunities, and student training
+\cite{bourke1998institutions,ben2016s}. However, interdisciplinary
+fields such as bioinformatics blur traditional boundaries by
+integrating data and methods from biology, computer science, and
+mathematics to address complex research challenges
 \cite{Ouzounis:2003,Eddy:2005,hogeweg2011roots}.
 
-Bioinformatics, is an interdisciplinary field that fuses biology,
-computer science, and mathematics, and now plays a pivotal role in
-modern biological research \cite{Ouzounis:2003}. The development of
-bioinformatic tools and software is critical for interpreting complex
-biological questions, such as what evolutionary, structural and
-functional analyses of genomic, transcriptomic, and proteomic data can
-tell us. The field of ``bioinformatics'' can include many overlapping
-research fields that include computational biology, biomathematics,
-biostatistics, medical informatics and other similar areas.
+Bioinformatics has become essential to modern biological research,
+facilitating evolutionary, structural, and functional analyses of
+genomic, transcriptomic, and proteomic data
+\cite{clement2018data,gauthier2019brief}. The development of accurate and
+scalable bioinformatic tools is critical for interpreting such large
+datasets. Which requires both biological insight and advanced
+computational skills to build algorithms. The advent of
+high-throughput technologies has driven the growth of bioinformatics,
+leading to the establishment of specialized groups within biology,
+computer science, and engineering faculties, each contributing to the
+field’s expansion
+\cite{hogeweg2011roots,gauthier2019brief}.
 
-Bioinformatics gained significant traction with the advent of
-high-throughput sequencing technologies, which created a need for
-robust computational tools to manage the large volumes of data
-generated \cite{Ouzounis:2003,hogeweg2011roots}. In response,
-departments specializing in bioinformatics began to emerge from
-biology, computer science, and engineering faculties, each bringing
-unique contributions to the field's development and expansion.
+The contributions of ``domain experts'' to bioinformatics from the
+biological and health sciences, such as genetics and molecular biology
+are essential as they ensure software tools are relevant and accurate.
+However, domain experts may lack the advanced computational expertise
+needed to develop sophisticated software. In contrast, fields like
+mathematics, engineering, and computational sciences — referred to
+here as ``development experts'' — offer expertise in algorithm
+development, mathematical modeling, statistics, and software
+engineering, essential for creating efficient and scalable
+bioinformatic tools.
 
-Bioinformatics integrates methods from various disciplines, fostering
-novel insights that are unattainable within a single field
-\cite{mazzocchi2019scientific}. This requires a deep understanding of
-biological sciences for interpreting data, and strong computational
-skills are required to develop algorithms and analyze the results at
-scale.
+It is possible that departmental differences may influence
+bioinformatic tool development by reflecting the distinct expertise,
+resources, and perspectives offered by different academic
+fields. Development experts excel in computational efficiency, while
+domain experts provide essential biological insights. Therefore, the
+success of a tool may depend on the integration of diverse
+skills rather than on the specific departmental affiliation of its
+developers.
 
-Experts in \textbf{biological and health sciences}, such as genetics
-and molecular biology, can ensure that software tools are biologically
-relevant and accurate by identifying critical questions and guiding
-computational tool development. However, biology departments may lack
-the advanced computational expertise required for developing
-sophisticated software. In our analysis, we have grouped the
-biological and health fields under the category of ``\textbf{domain
-  experts}''.
+The primary objective of this study is to examine whether the academic
+department affiliation of a corresponding author has a discernible
+outcome on the accuracy (i.e. correctness of predictions) of the
+bioinformatic tools they develop. Specifically, we aim to determine
+whether tools created by authors from domain-expert,
+development-expert of interdisciplinary fields differ in accuracy. To
+address this, we analyzed a benchmarked corpus of bioinformatic
+software tools and evaluated their accuracy based on the developers'
+academic affiliations.
 
-The \textbf{mathematical, engineering and computational sciences},
-referred to here as the ``\textbf{development experts}'', contribute
-significantly by bringing expertise in algorithm development,
-mathematical modelling, statistics, and software engineering
-principles. These skills are essential for creating efficient,
-scalable, and robust bioinformatic tools.  However, the challenge for
-development experts is gaining a deep understanding of biological domains,
-which is necessary to ensure software tools are both relevant
-and accurate.
-
-Departmental differences can influence the development of
-bioinformatic software tools because they reflect the varying
-expertise, resources, and perspectives that different academic fields
-offer. Development experts may excel in areas such as algorithm
-efficiency, mathematical modeling, data handling, and software
-engineering, while domain experts may bring a deep understanding of
-biological questions, data interpretation, limitations, and
-potentially better curation of control datasets. Consequently, the
-success of a tool may rely more on the integration of diverse
-forms of expertise rather than the specific departmental affiliation
-of its developers. This blending of skills can mitigate potential
-disparities between departments, resulting in comparable outcomes
-regardless of the department of origin.
-
-The primary objective of this study is to investigate whether the
-academic department affiliation of a corresponding author is
-associated with the accuracy of the bioinformatic software tools they
-develop. We aim to determine if tools developed by authors from
-certain academic fields, such as ``Bioinformatics'', ``Computer
-Science'' or ``Genetics'' show higher or lower accuracy compared to
-those from other fields. To achieve this, we analyzed a benchmarked
-corpus of bioinformatic software tools and evaluated their accuracy
-based on the academic affiliations of their developers.
-
-\begin{figure*}[ht!]
-\begin{center}
-  \includegraphics[width=0.6\textwidth]{upset-plots.pdf}
-\end{center}
+\begin{figure*}[t!]
+  \centering
+  \includegraphics[width=0.75\textwidth]{upset-plots.pdf}
   \caption{ The number and intersection of general (\textbf{A}) or
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section*{Results}
 
-We are interested in exploring the relationship between the accuracy
-of bioinformatic software tools and the academic fields of study of
-their developers. Using a published corpus of accuracy rankings
-for bioinformatic software tools, we mapped these tools to their
-respective academic fields and evaluated how software accuracy
-correlates with the developers' academic affiliations.
+We explored the relationship between the accuracy of bioinformatic
+software tools and the academic fields of their developers. Using a
+previously published corpus of benchmarked accuracy rankings
+\cite{gardner2024}, we mapped corresponding authors' addresses to
+standardized ``fields of study'' \cite{fields2014} and grouped them into
+broader categories.
 
-We used a published supplement \cite{gardner2024} to obtain previously
-benchmarked accuracy rankings and mapped corresponding authors'
-addresses to a standardized list of specific 'fields of study'
-\cite{fields2014}. These were further grouped into broader general
-fields and areas of expertise. Figure~\ref{fig:fig1} illustrates the
-number of tools representing each general and specific field ($N\ge 10$). Most
-bioinformatic software is developed by corresponding authors who list
-Genetics, Bioinformatics, Computer Science, or similar departments as
-their primary address (Figure~\ref{fig:fig1}A). Among the general
-fields, the Biological Sciences produce the most software tools,
-followed by the Computer Sciences (Figure~\ref{fig:fig1}B).
+%
+Figure~\ref{fig:fig1} shows the number of tools for the general and
+specific fields ($N\ge 10$). Most bioinformatic tools were developed
+by authors affiliated with Genetics, Bioinformatics, Computer Science,
+or similar departments. Among the general fields, Biological Sciences
+produced the most software tools, followed by Computer Sciences.
 
@@
 \begin{center}
-  \includegraphics[width=0.8\textwidth]{forest-z-Plot.pdf}
+  \includegraphics[width=0.65\textwidth]{forest-z-Plot.pdf}
 \end{center}
@@
 
-The mean proportion of wins (e.g., if tool `A' outranks tool `B', this
-counts as a win for tool `A') and the corresponding Z-scores provide a
-method to rank fields of study based on the relative accuracy of
-bioinformatic software benchmarked between 2005 and 2020, compared to
-the expected number of wins for random groupings (i.e., $wins =
-0.5$). A higher proportion of wins and a greater $(-1) * Z-score$
-indicate that tools from a particular department more frequently
-outperformed competing tools in independent benchmarks.
+We ranked fields based on the mean proportion of ``wins'' (i.e., for field '1' when its tool
+'A' outperforms tool 'B' in benchmark 'X') and calculated Z-scores
+to compare them against random expectations (i.e., $wins = 0.5$). A
+higher proportion of wins and lower $(-1) * Z-score$ indicate better
+overall accuracy.
 
-The specific and general field that outperformed all others is
-``Medical Informatics'', a branch of ``Technologies'', with 16
-software tools categorized under both fields. Among these, five are
-different parameter options for the MAFFT multiple sequence alignment
-tool \cite{katoh2008recent}.  These Medical Informatic tools have a mean proportion of wins of
-0.70 and 95\% confidence interval of $0.53-0.85$, which excludes the
-null value of 0.5. The corresponding Z-score is $-1.88$, and after correcting for multiple testing, the P-value is $0.29$.
+``Medical Informatics'', a branch of ``Technologies'', outperformed
+other fields, with a mean win proportion of 0.70 (95\% CI:
+$0.53-0.85$) and a Z-score of $-1.88$. However, P = 0.29 after
+multiple testing correction). Notably, this category includes five
+different parameter options for the MAFFT sequence alignment tool
+\cite{katoh2008recent}, and a further four separate corresponding
+authors that list either the Department or Center of Biomedical
+Informatics, Harvard Medical School as their affiliation
+\cite{kim2010rsw,yang2013diverse,kharchenko2014bayesian,ruan2020fast}. This
+and some other redundancies leaves just eight departments
+representing the medical informatics field here.
 
-At the other end of the spectrum is ``Bioinformatics''. Ironically,
-authors that list ``Bioinformatics'' in their address field tend to
-produce less accurate software for bioinformatic applications. The
-mean proportion of wins was 0.43, with a 95\% confidence interval of
-$0.33-0.53$. The corresponding Z-score is $1.20$ and after correcting
-for multiple testing the P-value is $0.46$.
+In contrast, ``Bioinformatics'' has the lowest rank, with a mean win
+proportion of 0.43 (95\% CI: $0.33-0.53$) and a Z-score of $1.20$ (P =
+0.46 after correction). Similarly, ``Engineering'' ranked low, with a
+win proportion of 0.34 (95\% CI: $0.16-0.60$) and a Z-score of $1.25$.
 
-The general field of ``Engineering'' also ranked low, with a mean
-proportion of wins of just 0.34 and a 95\% confidence interval of
-$0.16-0.60$. The corresponding Z-score is $1.25$. This general field
-comprises several smaller specific fields, such as "Bioengineering and
-Biomedical Engineering," "Computer Engineering," and "Electrical and
-Electronics Engineering." However, since none of these specific fields
-individually had more than ten corresponding software tools, they were
-excluded from the more detailed analysis.
+Other fields showed confidence intervals that included the null value
+of 0.5, with modest Z-scores ranging from -0.49 to 0.96, and P-values
+greater than 0.05.
 
-The remaining general and specific academic fields have confidence
-intervals that include the null value of 0.5, and relatively modest
-Z-scores that range from -0.49 to 0.96. The P-values for each were
-greater than $0.05$.
+When grouped by expertise type—software development experts,
+biological domain experts, or interdisciplinary experts—all categories
+had similar win proportions (0.51, 0.49, and 0.46,
+respectively). Interdisciplinary experts had the lowest Z-score of
+$-0.87$ (P = 0.46 after correction).
 
-For the highest level field classifications of software development
-expert, biological domain expert or interdisciplinary expert each had
-similar mean proportions of wins (0.51, 0.49 and 0.46
-respectively). The interdisciplinary experts had a lower Z-score of
-$-0.87$, which corresponds to $P=0.46$ after correcting for multiple
-testing.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 
-We tested the assumption that the speciality of academic departments
-reflect the quality of the research software they produce.  After
-correcting for multiple testing, we found no significant association
-between academic expertise type and the accuracy of bioinformatic
-tools, suggesting that academic department affiliation does not
-correlate with software quality. Similarly, both general and specific
-research fields showed no significant associations
-(Figure~\ref{fig:fig2}).
+We tested the assumption that academic department specialization
+reflects the quality of research software. After correcting for
+multiple testing, we found no significant association between academic
+expertise and the accuracy of bioinformatic tools. This suggests that
+department affiliation does not correlate with software quality, and
+neither general nor specific research fields showed any significant
+links to tool accuracy.
 
-Our earlier paper found that a long-term commitment to keeping
-software updated was the primary factor associated with accurate
-software tools. This current study complements that finding by demonstrating 
-that citation metrics (e.g. journal impact factors and
-author H-index), tool age, tool speed, and now academic fields of
-inquiry are \textbf{not} associated with software accuracy.
+A previous study indicated that maintaining software was the key
+factor in producing accurate tools, while citation metrics, tool age,
+speed are not associated with software accuracy \cite{Gardner:2022}.
+Our findings complement this by showing that academic field is also not
+associated with software accuracy.
 
-We focus on software tool accuracy here \cite{weber2019essential}. While speed, usability and
-some features of software tools are important, in our opinion the
-primary concern for bioinformatic software tools is the accuracy of
-the results they produce. As poor predictions may have long-term consequences for
-our general research field.
+While other aspects of bioinformatic tools, such as speed and
+usability, are important, we emphasize that accuracy should remain the
+top priority, as poor predictions can have long-term impacts on
+research \cite{weber2019essential}.
 
-Medical Informatics, under the broader category of "Technologies," is
-identified as the top-performing group in developing accurate
-bioinformatic software tools. The tools include a number of methods
-for structural variation detection, single-cell profiling, long-read
-assembly, multiple sequence alignment and are derived from several
-different research teams.
+Medical Informatics was the top-performing field in developing
+accurate tools, these include methods for structural variation
+detection, single-cell profiling, long-read assembly, multiple
+sequence alignment and are derived from a limited number of research
+teams. However, tools from Bioinformatics and Engineering
+ranked lower, though these differences were not statistically
+significant.
 
-Bioinformatics and Engineering ranked lower in terms of software
-accuracy. Tools developed by authors who affiliated with
-"Bioinformatics" typically had slightly lower accuracy than that of
-other fields. However, this was not a statistically significant
-finding. In addition (further bad news for this author),
-``Biochemistry'' was similarly ranked, again this was not a
-statistically significant finding.
+Therefore, an individual’s department is not a reliable indicator of the
+quality of the software they produce. Academic affiliation should not
+be used as a proxy for assessing the potential success of software
+development projects.
 
-This leads us to conclude that an individual's host department are not
-reflective of the quality of software that they are capable of
-producing ($p>0.05$ in all instances). As a consequence, the academic
-department should not be used as a proxy for judging the potential of
-software development projects.
+\textbf{Limitations:} Some benchmarks include multiple tool options,
+potentially introducing non-independent effects. The accuracy metrics
+are diverse, with some limitations (e.g., issues with "accuracy" in
+class-imbalanced datasets \cite{luque2019impact} and criticisms of the
+N50 metric for sequence assembly \cite{xie2021pdr}). Additionally,
+smaller benchmarks and smaller fields may exaggerate rank shifts. We
+mitigate this in part by only considering departments with 10 or more
+corresponding tools.
 
-\textbf{Limitations:} Some benchmarks rank multiple tool options,
-which can introduce modest to large effects as these options may not
-be independent. Additionally, the accuracy metrics used are diverse,
-and some may be flawed under certain conditions; for example,
-``accuracy'' can be misleading with large class-imbalances
-\cite{luque2019impact}, and the N50 metric for sequence assembly has
-been criticized by some commentators \cite{xie2021pdr}. Moreover, some
-benchmarks are relatively small, so even small changes in rank can
-have a substantial impact on the proportion of wins for intermediate
-rankings. Lastly, the cohort of benchmarks has not been updated to
-include more recent results.
+%Also, our data set has not been updated with more recent
+%tools.
 
-There is often a disconnect between a developer’s training and their
-departmental affiliation, as evidenced by this author whose diverse
-academic and professional background spans mathematics,
-bioinformatics, computer science, molecular biology, and genomics, yet
-currently lists their affiliation with a Biochemistry Department. The
-``Biochemistry'' departmental label does not accurately reflect the
-author's training or recent publications. In fact, the author admits
-to not recalling key details of glycolysis, the citric acid cycle, or
-the structures of amino acids and proteins.
+There may be little connection between a researcher’s training and
+their listed department, as illustrated by this author’s background
+which began in mathematics, before taking positions in departments of
+computer science, molecular biology and bioinformatics, and now is
+affiliated with a Biochemistry Department.
 
-The last, or corresponding, author is typically the principal
-investigator who leads a project. They may have limited direct
-involvement in the development of any software tool, while
-their role is mainly to provide resources and overall direction for
-the project. However, there is likely to be a significant overlap between the
-department of the first author, who is usually the primary developer
-of the tool, and the department of the last author (though this was
-not tested in this study). Therefore, we expect that the results will
-be broadly similar if first-author departments were analyzed instead.
+The corresponding (last) author, is typically the principal investigator and
+may not be the primary tool developer, but rather oversees the
+project. There is likely to be a strong overlap between the departments of the first
+and last authors, but this was not explored in the current study.
 
-In conclusion, this study does not find strong evidence that the
-academic department affiliation of authors is associated with the
-accuracy of bioinformatic software tools. This challenges the
-assumption that departmental expertise correlates with
-software quality. Future research could explore other factors, such
-as the nature of interdisciplinary collaborations or the training
-of developers, to better understand what factors contribute to
-the development of high-quality bioinformatic tools.
+\textbf{Final words:} This study does not find strong evidence linking
+academic department affiliation with bioinformatic software
+accuracy ($p>0.05$ in all instances). Future research should investigate other factors, such as
+interdisciplinary collaborations and developer training, to understand
+what drives high-quality tool development. Addressing potential biases
+against interdisciplinary work \cite{bromham2016interdisciplinary} and ensuring long-term support for
+essential software infrastructure will also be critical for advancing the
+field \cite{siepel2019challenges}.
 
-%Open questions: what impact does training have on improving software quality?
-%We could end with echoing the call for more training and a continued 
-%e.g. \cite{johanson2018software,siepel2019challenges}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-\section*{Methods}
+\subsection*{Methods}
 
-~~~~\textbf{Pre-registration:} This study's desired sample size, included variables, hypotheses, and
-planned analyses were pre-registered on the Open Science Framework
+The data, scripts, figures and manuscript draft files are availble at the GitHub repository:\\
+\href{https://github.com/ppgardne/departments-software-accuracy}{https://github.com/ppgardne/departments-software-accuracy}
+
+~~~~\textbf{Pre-registration:} The methods for this study followed the pre-registered proposal outlined
+prior to any unpublished data collection \cite{gardner2024}.
 %(https://doi.org/10.17605/OSF.IO/92PTZ)
-prior to any unpublished data being collected \cite{gardner2024}.
 
@@
 
-%The methodology section should provide more detail about how the software tools were selected and mapped to academic fields.
-It’s important to describe how biases or inconsistencies in mapping were addressed.
+%The methodology section should provide more detail about how the
+%software tools were selected and mapped to academic fields. It’s
+%important to describe how biases or inconsistencies in mapping were
+%addressed.
+
 \textbf{Mapping tools to academic field:} For each software tool, the
@@
 ranged from 3 to 50. In order to obtain a representative measure of
-accuracy for a field that accounts for the diversity in accuracy measures and
-number of ranked tools, we employed a rank-based and
-bootstrapping strategy.
-We randomly sampled, with replacement, sets of 200 tools from the total of 498 tools.
-For each tool, a corresponding benchmark was selected at random, and the number of times the tool ``won'' against another tool was recorded, along with the total number of pairwise comparisons made.
-These counts of wins and total comparisons were then assigned to the corresponding specific, general, and expertise areas.
-This process was repeated 1,000 times to estimate the mean proportions of wins for each field, along with a 95\% confidence interval for these values (Figure~\ref{fig:fig2}A).
-Additionally, we calculated a Z-score for each field to determine the number of standard deviations the mean number of wins deviates from the expected null value of 0.5 for randomly grouped tools
-(Figure~\ref{fig:fig2}B).
+accuracy for a field that accounts for the diversity in accuracy
+measures and number of ranked tools, we employed a rank-based and
+bootstrapping strategy.  We randomly sampled, with replacement, sets
+of 200 tools from the total of 498 tools. For each tool, a
+corresponding benchmark was selected at random, and the number of
+times the tool ``won'' against another tool was recorded, along with
+the total number of pairwise comparisons made. These counts of wins
+and total comparisons were then assigned to the corresponding
+specific and general departments, and expertise areas.
+In other words, a tool ranked second in a benchmark of 11 tools will
+contribute 9 wins and 10 comparisons to the totals for its
+corresponding fields.
+
+This process was repeated
+1,000 times to estimate the mean proportions of wins for each field,
+along with a 95\% confidence interval for these values (Figure~2A).
+Additionally, we calculated a Z-score for each field to determine the
+number of standard deviations the mean number of wins deviates from
+the expected null value of 0.5 for randomly grouped tools (Figure~2B).
 
@@
 
-\noindent\textbf{Data and analysis scripts availability}
+%% In brief, software ranks from bioinformatic software benchmarks were
+%% gathgered \cite{Gardner:2022}, these cover 68 publications ranking the
+%% accuracy 498 distinct software tools in total. Software tools were
+%% mapped to departmental field using the address(es) of corresponding
+%% authors, these were further mapped to the closest associated ``fields
+%% of study'' defined by the National Science Foundation
+%% \cite{fields2014}. We analysed fields at three hierarchical levels:
+%% \textbf{1.} specific fields (e.g. ``genetics'', ``computer science'',
+%% ``bioinformatics''), \textbf{2.} general fields (e.g. ``biological
+%% sciences'', ``computer sciences''), \textbf{3.}  expertise area
+%% development experts, domain experts or interdisciplinary experts.
 
-The data, scripts, figures and manuscript draft files are availble at the GitHub repository:\\
-\href{https://github.com/ppgardne/departments-software-accuracy}{https://github.com/ppgardne/departments-software-accuracy}
+%% We then employed a rank-based bootstrapping strategy.  Sets of 200
+%% tools are randomly sampled, with replacement, a corresponding
+%% benchmark was selected at random for each tool, and the number of
+%% times the tool ``won'' against another tool was recorded, along with
+%% the total number of pairwise comparisons made. These counts of wins
+%% and total comparisons were then assigned to the corresponding
+%% specific, general, and expertise areas. This was repeated 1,000 times
+%% to estimate the mean proportions of field wins and a 95\% confidence
+%% interval (Figure~\ref{fig:fig2}A). Corresponding
+%% Z-scores indicate the number of standard deviations
+%% from the expected null of 0.5 each field lies (Figure~\ref{fig:fig2}B).
 
-\noindent\textbf{Funding:} This research was supported by the MBIE data
+%% $Z-score=\frac{0.5-\mu}{\sigma}$
+
+%% Where $\mu$ is the mean, $\sigma$ is the standard deviation, $x$ is the raw
+%% value. P-values are computed from the absolute value of the Z-scores,
+%% $P[X > x]$. P-values are corrected for multiple testing by
+%% controlling the false discovery rate method
+%% \cite{benjamini1995controlling}.
+
+\subsection*{Acknowledgements}
+
+  This research was supported by the MBIE data
 science platform ``Beyond Prediction: explanatory and transparent data
@@
 
-\noindent\textbf{Authors' contributions:} this work was conceived by 
-PPG, designed by PPG, the analysis was carried out by PPG, interpretation of the results was by PPG,
-the manuscript was drafted by  PPG.
-All authors revised and edited the final manuscript. All authors approve the submitted version.
+Associate Professor Anthony Gitter (Department of Biostatistics and
+Medical Informatics, University of Wisconsin-Madison) for pointing out
+the over-representation of Harvard Medical School affiliations in the
+Medical Informatics catagory.
 
-%----------------------------------------------------------------------------------------
-%       REFERENCE LIST
-%----------------------------------------------------------------------------------------
+% Bibliography
 \bibliographystyle{unsrt}
 \bibliography{references.bib}
+%pdflatex manuscript.tex && bibtex  ./manuscript  && pdflatex manuscript.tex && pdflatex manuscript.tex
 

### 2024-11-14 docs/manuscript.tex
Updated manuscript with reviewer suggestions.

@@
 \definecolor{color3}{RGB}{5,5,5} % Color of the boxes behind the abstract and headings
+
+%\definecolor{rv}{RGB}{0,0,200} %colour highlighting revisions
+\definecolor{rv}{RGB}{0,0,0} %colour masking revisions
 
@@
 
+\textcolor{rv}{The accuracy of bioinformatic software tools are of
+  critical importance to the field \cite{weber2019essential}. Yet
+  several factors can influence accuracy assessments. Previous work
+  has shown that author conflicts can bias tool accuracy assessments
+  upwards \cite{buchka2021optimistic}. Conversely, commonly assumed
+  features of accurate software such as author reputation
+  (e.g. H-index), tool citation rates, journal impact, runtime, and
+  age of tools had no significant association with software accuracy
+  \cite{Gardner:2022}.  Instead, features illustrating sustained
+  development of tools, such as the number of releases, and activity
+  on a popular distributed version control website was significantly
+  associated with accuracy \cite{Gardner:2022}.}
+
 The primary objective of this study is to examine whether the academic
@@
 whether tools created by authors from domain-expert,
-development-expert of interdisciplinary fields differ in accuracy. To
-address this, we analyzed a benchmarked corpus of bioinformatic
-software tools and evaluated their accuracy based on the developers'
-academic affiliations.
+development-expert of interdisciplinary fields differ in accuracy.
+\textcolor{rv}{This question was motivated by the possibility
+  that certain grant reviewers and other research assessors may base
+  their perception of expertise on departmental affiliation rather
+  than more pertinent information such as research experience.} To
+address this challenge we have analyzed a published corpus of
+benchmarked bioinformatic tools and evaluated relative accuracy ranks
+based on the developers' academic affiliations.
 
@@
   \centering
-  \includegraphics[width=0.75\textwidth]{upset-plots.pdf}
-  \caption{ The number and intersection of general (\textbf{A}) or
-  specific (\textbf{B}) fields that have contributed to bioinformatic
-  software tools included in recent benchmark studies.  The number and
-  intersection of specific fields.  }
+  \includegraphics[width=0.75\textwidth]{figure1.pdf}
+  \caption{ \textcolor{rv}{ The figure displays the number and
+      intersection of bioinformatic tools classified by general fields (Panel \textbf{A})
+      and specific fields (Panel \textbf{B}), as listed by the corresponding
+      authors who published software included in benchmark
+      studies. Panel \textbf{C} presents the classification of general and
+      specific fields of study used by the ``National Science Foundation''
+      relevant to this study. On the right, it shows the additional
+      expertise categories that we introduced for this analysis.
+ }  }
 \label{fig:fig1}
@@
 \cite{gardner2024}, we mapped corresponding authors' addresses to
-standardized ``fields of study'' \cite{fields2014} and grouped them into
-broader categories.
+standardized ``fields of study'' \cite{fields2014} and grouped them
+into broader categories. Fields of study is a hierarchical
+classification that includes specific fields (e.g. ``Genetics'' or
+``Mathematics'') or general fields (e.g. ``Biological Sciences'' or
+``Mathematics and Statistics''). We further classified these into
+expertise areas of ``Development'' (e.g. Computer Science), ``Domain''
+(e.g. Biological Sciences) and ``Interdisciplinary''
+(e.g. Bioinformatics or listing both a Biological and Computational
+Sciences affiliation).
 
 %
-Figure~\ref{fig:fig1} shows the number of tools for the general and
-specific fields ($N\ge 10$). Most bioinformatic tools were developed
-by authors affiliated with Genetics, Bioinformatics, Computer Science,
-or similar departments. Among the general fields, Biological Sciences
-produced the most software tools, followed by Computer Sciences.
+\textcolor{rv}{Figure~\ref{fig:fig1} shows the number of tools (set/intersection sizes) for the general
+and specific fields (for sets sizes $\ge 10$), as well as the intersections between
+fields due to corresponding authors with multiple
+affiliations. E.g. 11 tools have been published by authors who list
+both a Genetics and Computer Science department as their
+affiliation.} Most bioinformatic tools were developed by authors
+affiliated with Genetics, Bioinformatics, Computer Science, or similar
+departments. Among the general fields, Biological Sciences produced
+the most software tools, followed by Computer Sciences.
 
@@
 \begin{center}
-  \includegraphics[width=0.65\textwidth]{forest-z-Plot.pdf}
+  \includegraphics[width=0.65\textwidth]{figure2.pdf}
 \end{center}
@@
 benchmarks are publically available \cite{Gardner:2022}, these include data from
-68 publications that rank the accuracy of different sets of 498
-distinct software tools.
+68 publications, \textcolor{rv}{128 benchmark rankings of} different sets of 498
+distinct software tools (See Supplementary Table 1).
+
+\textcolor{rv}{In brief, the criteria for inclusion of a
+  benchmark study are based on the definition of ``neutral comparison
+  studies'' \cite{boulesteix2013plea}: 1. The main focus of the
+  article is the comparison and not the introduction of a new
+  tool. 2. The authors should be neutral (i.e. not tool
+  developers). 3. The test data and evaluation criteria should be
+  sensible.}
 
@@
 
+\textcolor{rv}{The author profusely thanks the recent anonymous grant
+  reviewers who kindly noted the author’s
+  bioinformatics/machine-learning skills as 'weak' or 'lacking
+  expertise'. We assume they were swayed more by our affiliation than
+  by minor details like... a publication record. This charming
+  oversight, of course, inspired the paper you're reading now.}
+
 % Bibliography

### 2024-11-14 bin/plotResults.R

@@ -86,3 +86,21 @@ UpSet(sm,
       )
 dev.off()
 
+######################################################################
+#Evaluate confounding by age:
+#Run: ../bin/extractAgeData.pl
+
+yearPublished <- read.table("yearPublishedByField.tsv", header=T, sep="\t")
+medYear<-median(yearPublished$yearPublished)
+pdf(file=    "../docs/figures/toolAgesByField.pdf", width = 24,  height = 15)
+par(mfrow = c(1,2), las=2, mar=c(5, 10, 4, 2) + 0.1, cex=2)
+bymedian <- with(yearPublished, reorder(generalField, yearPublished, median))
+#boxplot( yearPublished ~ generalField,  data = yearPublished, col = "blue",ylab='',horizontal=TRUE)
+boxplot( yearPublished ~ bymedian,  data = yearPublished, col = "blue",ylab='',horizontal=TRUE)
+lines(c(medYear, medYear), c(0,5), lwd=2, lty=2, col='red')
+bymedian <- with(yearPublished, reorder(specificField, yearPublished, median))
+boxplot( yearPublished ~ bymedian, data = yearPublished, col = "blue",ylab='',horizontal=TRUE)
+lines(c(medYear, medYear), c(0,15), lwd=2, lty=2, col='red')
+dev.off()
+
+
